thesis

### 2025-01-02 thesis.qmd
morphology thesis

---
title: "thesis"
author: "Dimitris Papachristopoulos"
format: 
  pdf: default
  #logo: ./figure/
  arxiv-pdf:
    keep-tex: true    
toc: true
toc-depth: 3
theme: minti
bibliography: ./My_Library.bib
execute: 
  echo: false
  eval: true
  warning: false
editor: visual
---

@@

@haslbauer2023 argues that the Delayed-$\tau$ model is opposed to the Lilly-Madau plot ([@madau2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses [^1] from the UNGC catalog (@karachentsev2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LCV), allows the investigation of the SFR throughout the life each galaxy and so we can find the expected time of peak of the SFR.

@@

@@

### Hubble-de Vaucouleurs Classification

@@

SFR can be estimated using various photometric or spectroscopic methods based on the luminosity of at least one spectral band or the intensity of a spectral line. Different luminosities and intensities trace distinct emission mechanisms, offering insights into a galaxy’s radiation sources. Below are common methods:

[^2]

[^2]:
    -   [@calzettiStarFormationRate2012], [@ASTR620GalaxiesFall]

@@

According to [@madau2014], this method often underestimates the SFR, since different galaxy populations may systematically follow distinct absorption mechanisms depending on their characteristics.

@@

Parameterized SFH models are commonly used, offering simplicity through a few parameters (@carnallHowMeasureGalaxy2019a):

@@

```{bash}
gnuplot -e "
set terminal svg;
set output 'figure/combined_models.svg';
set xlabel 'Time (t)';
set ylabel 'SFR(t)';
set title 'Star Formation Rate Models';
set grid;
plot exp(-x/8) title 'Exponential Decline (tau=2)' with lines lw 2, \
     x*exp(-x/2) title 'Delayed Exponential (tau=2)' with lines lw 2, \
     (1/1)*exp(-((log(x)-2)**2)/(2*1**2)) title 'Log-Normal (tau=1, T0=2)' with lines lw 2, \
     1 / ((x/2)**2 + (x/2)**3) title 'Double Power Law (alpha=2, beta=3)' with lines lw 2;
"
```

```{r}
#| fig.cap="Star Formation Rate over the time of star formation, for different parametric models"
library(ggplot2)
library(dplyr)

# Define time values (positive times)
t <- seq(0.1, 30, length.out = 300)

# Generate data for each model
data <- data.frame(
  t = rep(t, 4),
  SFR = c(
    exponential_decline(t, tau = 2),
    delayed_exponential(t, tau = 2),
    log_normal(t, tau = 2, T0 = 12),
    double_power_law(t, tau = 2, alpha = 2, beta = 8)
  ),
  Model = rep(c("Exponential Decline", "Delayed Exponential", "Log-Normal", "Double Power Law"), each = length(t))
)

# Plot all models
ggplot(data, aes(x = t, y = SFR, color = Model)) +
  geom_line(size = 1) +
  labs(
    title = "Star Formation Rate Models",
    x = expression("Star Formation time (" * t[sf] * ") [Gyr]"),
    y = expression("SFR(" * t[sf] * ") ["* M['☉']/yr *"]" ),
    color = "Model"
  ) +
  scale_x_continuous(limits = c(0, 30)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

@@

The Lilly-Madau plot is one of the most important plots in the field of galaxy evolution. It describes how the SFRD of the universe evolved over time, with observational data. to figure out

@@

$V = H_0\times d$, where $H_0 \approx 69.8 \text{ km/s/Mpc}$ is the Hubble constant and $d$ is the distance between the two galaxies.[^3]

[^3]: We also have non-Hubble motions $V = H_0\times d + V_0$, where $V_0$ is the peculiar velocity and it could, for example, be due to galaxy cluster dynamics. For the current explanation we are going to ignore it. This way the radial velocity $v$ is equal to $V$

@@

>>>>

---
title: "thesis"
author: "Dimitris Papachristopoulos"
format: 
  pdf: default
  #logo: ./figure/
  arxiv-pdf:
    keep-tex: true    
    link-citations: true
    link-bibliography: true
toc: true
toc-depth: 3
theme: minti
bibliography: ./My_Library.bib
execute: 
  echo: false
  eval: true
  warning: false
editor: visual
---

@@

@haslbauerCosmologicalStarFormation2023 argue that the Delayed-$\tau$ model is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses [^1] from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LCV), allows the investigation of the SFR throughout the life each galaxy and so we can find the expected time of peak of the SFR.

@@

This thesis will focus on how a specific parametric model tries to explain the Star-Formation Histories of galaxies, but to do that, we first need to understand what a galaxy is, what it is made of, the mechanisms of its evolution, and how we can distinguish them.

The study of galaxies is a very active field of astronomy since it is a relatively young discipline. Until 1920, astronomers who observed spiral nebulae were not certain what they were. In 1921, two papers were published: one argued that the Milky Way constituted the whole Universe and that the spiral nebulae were part of it, while the second argued that each spiral nebula was, in fact, a distinct "island universe" [@shapleyScaleUniverse1921]. This debate became known as the Great Debate[^2].

[^2]: @hoskinGreatDebateWhat1976

In 1925, Edwin Hubble put an end to the debate by showing that the distances to the spiral nebulae were far too great compared to other objects within the Milky Way. The method he used involved observing Cepheid variable stars in these nebulae. By applying the period-luminosity relationship, which had been discovered by Henrietta Swan Leavitt [@leavitt1777VariablesMagellanic1907], Hubble was able to determine their distances. This groundbreaking discovery confirmed that spiral nebulae were indeed separate galaxies, marking the beginning of extragalactic astronomy and revolutionizing our understanding of the Universe.

Today we know that galaxies are large-scale structures containing, Stars, gas and dust, Stellar and planetary systems and Stellar remnants (white dwarves, neutron stars and black holes). Those structures are held together by their gravity, having Stellar Masses more than $\sim 10^5\ M_\odot$, and an average diameter of $\sim 4\ \text{kpc}$. They are extremely diverse systems, and each galaxy differs in mass, size, brightness, stellar populations and morphology. This is exactly why we have created systems to classify them.

@@

The most common method of classification is the *Hubble-de Vacouleur* classification, which categorizes galaxies based on their morphology.

-   **Elliptical Galaxies (E):** Ellipsoidal shapes with smooth brightness profiles, containing older stars and minimal interstellar matter. Their eccentricity takes values from 0 to 0.7, so we can categorize them even further, form E0 to E7

-   **Lenticular Galaxies (S0):** Intermediate between elliptical and spiral galaxies,

-   **Spiral Galaxies:** Characterized by a central bulge and spiral arms. Depending on the arm tightness we can categorize them as *a,b,c,d* from tight to looser, and depending if they have a bar or not they are subdivided into:

    -   **Unbarred Spirals (S):** No central bar; classified as Sa, Sb, Sc, etc., based on arm tightness and bulge size.

    -   **Barred Spirals (SB):** Feature a central bar; denoted as SBa, SBb, SBc, etc.

-   **Irregular Galaxies (Irr):** Lack regular structure, often rich in gas and dust with high star formation rates.

Intermediate galaxies between ellipticals and spirals are called **Lenticular galaxies (S0-SB0)**, featuring a central bulge and disk but lacking significant spiral structure.

[![Tuning-fork-style diagram of the Hubble sequenceBy Cosmogoblin - Own work, CC0, https://commons.wikimedia.org/w/index.php?curid=121743256](figure/1311px-Hubble_Tuning_Fork_diagram.svg.png){fig-align="center" width="400"}](https://commons.wikimedia.org/w/index.php?curid=121743256)

| **Hubble** | E | E-S0 | S0 | S0/a | Sa | Sa-b | Sb | Sb-c | Sc | Sc-Irr | Irr |
|:----:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|
| **T** | -5 | \[-4,-3\] | \[-2,-1\] | 0 | 1 | 2 | 3 | 4 | \[5,6,7\] | 8 | \[9,10,11\] |

: This table was based on the \LaTeX files of [@RC3ThirdReference]

@@

SFR can be estimated using various photometric or spectroscopic methods based on the luminosity of at least one spectral band or the intensity of a spectral line. Different luminosities and intensities trace distinct emission mechanisms, offering insights into a galaxy’s radiation sources. Below are common methods:[^3]

[^3]: @calzettiStarFormationRate2012, @mushotzkyASTR620GalaxiesFall2017

@@

According to [@madauCosmicStarFormation2014], this method often underestimates the SFR, since different galaxy populations may systematically follow distinct absorption mechanisms depending on their characteristics.

@@

```{r}
#| fig.cap="Star Formation Rate over the time of star formation, for different parametric models"
library(ggplot2)
library(dplyr)
library(latex2exp)

# Define time values (positive times)
t <- seq(0.01, 12, length.out = 200)

# Generate data for each model
data <- data.frame(
  t = rep(t, 4),
  SFR = c(
    exponential_decline(t, tau = 2),
    delayed_exponential(t, tau = 2),
    log_normal(t, tau = 2, T0 = 10),
    double_power_law(t, tau = 2, alpha = 2, beta = 5)
  ),
  Model = rep(c("\nExponential Decline\n(τ=2 Gyr)", "\nDelayed Exponential\n(τ=2 Gyr)", "\nLog-Normal\n(τ = 2 Gyr, To = 10 Gyr)", "\nDouble Power Law\n(τ = 2 Gyr, α = 2, β = 5)"), each = length(t))
)

# Plot all models
ggplot(data, aes(x = t, y = SFR, color = Model)) +
  geom_line(size = 1) +
  labs(
    title = "Star Formation Rate Models",
    x = TeX("Star Formation time ( $t_{sf}$ ) [Gyr]"),
    y = TeX("SFR($t_{sf}$) [ M$_*$/yr ]" ),
    color = "Model"
  ) +
  scale_x_continuous(limits = c(0, 12)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

Parameterized SFH models are commonly used, offering simplicity through a few parameters [@carnallHowMeasureGalaxy2019a]:

@@

@@

The Lilly-Madau plot is one of the most important plots in the field of galaxy evolution. It describes how the SFRD of the universe evolved over time, with observational data. But to understand it, we first need to understand how the observed age of the Universe and the redshifts of galaxies are related.

@@

$V = H_0\times d$, where $H_0 \approx 69.8 \text{ km/s/Mpc}$ is the Hubble constant and $d$ is the distance between the two galaxies.[^4]

[^4]: We also have non-Hubble motions $V = H_0\times d + V_0$, where $V_0$ is the peculiar velocity and it could, for example, be due to galaxy cluster dynamics. For the current explanation we are going to ignore it. This way the radial velocity $v$ is equal to $V$

@@

## Lilly-Madau Plot
